---
output: html_document
editor_options: 
  chunk_output_type: console
---
-------------------------------------------------------------------------------------------------------------------------------------
Individual quality and environmental factors interact to shape reproduction and survival in a resident bird of prey

Roman Bühler, Thomas V. Riecke, Kim Schalcher, Alexandre Roulin and Bettina Almasi



-------------------------------------------------------------------------------------------------------------------------------------

The complete code is written in R Version: 4.1.2

ATTENTION: Be aware that changes to the JAGS algorithm between versions and the simulation of the posterior distribution can 
lead to slight changes in the estimated coefficients, but they do not change the obtained results described in the publication.

 

load packages and data and scale mat function

```{r}
rm(list=ls())

library(jagsUI)
library(vioplot)
library(reshape2)
library(RColorBrewer)
library(tictoc)
library(tidyverse)
library(gridExtra)
library(ggforce)
library(beepr)
library(matrixStats)
library(scales)

load("Dataset Buehler et al. 2023.Rdata")

#function to scale over the whole matrix and not only column wise
scale.mat<-function(mat){
  sd=sd(mat, na.rm=T)
  mean=mean(mat, na.rm=T)
  mat<-(mat-mean)/sd
}

```

Explanation of the matrices in the dataset

alive: alive matrix, 1 if found alive, 0 if not seen alive
dead: dead matrix, 1 if found dead, 0 if not found dead
sex: 1=female, 2=male
gps: equipped=2, not equipped=1
age.cat: 1=yearling, 2=older
egg2: number of total eggs produced during the season
nestling2: total number of nestlings per season
fledge2: number of total fledglings per season
col: coloration from white to brown
col.p: diameter of black points on plumage
cort.b: baseline corticosteroid
cort.r: response corticosteroid
lay: laying date, numeric, origin="1970-01-01"
BCIe: Body condition index during egg laying period
BCIn: Body condition index during brood rearing period


```{r, eval=F}

save(alive,dead,sex,gps,age.cat,egg2,nestling2,fledge2,col,col.p,cort.b,cort.r,lay,BCIe,BCIn,preytot.lay.pl,preytot.play.pl,preytot.nestling.pl,preytot.w1.pl,preytot.lay.tr,preytot.play.tr,preytot.nestling.tr,preytot.w1.tr,file="Dataset Buehler et al. 2023.Rdata")




```



data formatting

cleanup
```{r}
n.years <- ncol(alive)

#eliminate juvenile encounters/releases

alive[age.cat == 0] <- 0
table(age.cat,alive)


##get first alive encounters

get.f <- function(x){min(which(x == 1))}
f <- apply(alive, 1, get.f)


```



as we need at least one coloration value to extrapolate we attribute the mean of each individual to the first encounter, if the bird was never measured we attribute the sex specific mean over all years

```{r}

col.init<-col

#for col
for (i in 1:nrow(col)){
  if(!is.na(col[i,f[i]])) next
  rowmean<-sum(col.init[i,], na.rm=T)/length(which(col.init[i,]>0))
  if(is.na(rowmean)& sex[i]==1){col[i,f[i]]<-mean(col.init[which(sex==1),], na.rm=T)}
  if(is.na(rowmean)& sex[i]==2){col[i,f[i]]<-mean(col.init[which(sex==2),], na.rm=T)}
  if(!is.na(rowmean)){col[i,f[i]]<-rowmean}
}
col<-round(col, digits=2)

#for col.p

col.p.init<-col.p

for (i in 1:nrow(col.p)){
  if(!is.na(col.p[i,f[i]])) next
  rowmean<-sum(col.p.init[i,], na.rm=T)/length(which(col.p.init[i,]>0))
  if(is.na(rowmean)& sex[i]==1){col.p[i,f[i]]<-mean(col.p.init[which(sex==1),], na.rm=T)}
  if(is.na(rowmean)& sex[i]==2){col.p[i,f[i]]<-mean(col.p.init[which(sex==2),], na.rm=T)}
  if(!is.na(rowmean)){col.p[i,f[i]]<-rowmean}
}
col.p<-round(col.p, digits=2)
```

we need to do the same for corticosterone but here we take year specific mean

```{r}
#for cort.b

cort.b.init<-cort.b


for (i in 1:nrow(cort.b)){
  if(!is.na(cort.b[i,f[i]])) next
  rowmean<-sum(cort.b.init[i,], na.rm=T)/length(which(cort.b.init[i,]>0))
  if(is.na(rowmean)& sex[i]==1){cort.b[i,f[i]]<-mean(cort.b.init[,f[i]], na.rm=T)}
  if(is.na(rowmean)& sex[i]==2){cort.b[i,f[i]]<-mean(cort.b.init[,f[i]], na.rm=T)}
  if(!is.na(rowmean)){cort.b[i,f[i]]<-rowmean}
}
cort.b<-round(cort.b, digits=2)




#for cort.r

cort.r.init<-cort.r

for (i in 1:nrow(cort.r)){
  if(!is.na(cort.r[i,f[i]])) next
  rowmean<-sum(cort.r.init[i,], na.rm=T)/length(which(cort.r.init[i,]>0))
  if(is.na(rowmean)& sex[i]==1){cort.r[i,f[i]]<-mean(cort.r.init[,f[i]], na.rm=T)}
  if(is.na(rowmean)& sex[i]==2){cort.r[i,f[i]]<-mean(cort.r.init[,f[i]], na.rm=T)}
  if(!is.na(rowmean)){cort.r[i,f[i]]<-rowmean}
}
cort.r<-round(cort.r, digits=2)

```


make gps mat a 1/2 mat instead of 0/1

```{r}
gps<-gps+1
```

make age 0/1 instead of 1/2
```{r}
age.cat<-age.cat-1

age.cat[age.cat<0]<-NA
```

get dead recoveries

```{r}
d <- apply(dead, 1, get.f)
d[d == Inf] <- NA
table(f,d)

```

get last alive encounter

```{r}
get.l <- function(x){max(which(x == 1))}
l <- apply(alive, 1, get.l)
```

fill in known latent states

```{r}
z <- matrix(NA, nrow(alive), ncol(alive))
for (i in 1:nrow(z)){
  if (l[i] > f[i]){
    z[i, f[i]:l[i]] <- 1
  }
  if (!is.na(d[i])){
    z[i, f[i]:(d[i]-1)] <- 1
  }
  z[i,f[i]] <- NA
}
```

calculate laying date as days from 01.01 of every year
```{r}
lay.d<-lay

as.Date(17167, origin="1970-01-01")
as.Date(17532, origin="1970-01-01")
as.Date(17897, origin="1970-01-01")
as.Date(18262, origin="1970-01-01")
as.Date(18628, origin="1970-01-01")


lay.d[,1]<-lay.d[,1]-17167
lay.d[,2]<-lay.d[,2]-17532
lay.d[,3]<-lay.d[,3]-17897
lay.d[,4]<-lay.d[,4]-18262
lay.d[,5]<-lay.d[,5]-18628

```


We do not predict for the fifth year and thus need to remove the values from the dataset, otherwise we will include them while scaling
```{r}
BCIe[,5]<- NA
BCIn[,5]<- NA
cort.b[,5]<- NA
cort.r[,5]<- NA
col.p[,5]<- NA
col[,5]<- NA
egg2[,5]<- NA
fledge2[,5]<- NA
nestling2[,5]<- NA
lay.d[,5]<- NA
```

scale matrices
```{r}
BCIe.x<-scale.mat(BCIe)
BCIn.x<-scale.mat(BCIn)
cort.b.x<-scale.mat(cort.b)
cort.r.x<-scale.mat(cort.r)
col.x<-scale.mat(col)
col.p.x<-scale.mat(col.p)
lay.d.x<-scale.mat(lay.d)



preytot.lay.pl.x<-scale.mat(preytot.lay.pl)
preytot.play.pl.x<-scale.mat(preytot.play.pl)
preytot.nestling.pl.x<-scale.mat(preytot.nestling.pl)
preytot.w1.pl.x<-scale.mat(preytot.w1.pl)
preytot.lay.tr.x<-scale.mat(preytot.lay.tr)
preytot.play.tr.x<-scale.mat(preytot.play.tr)
preytot.nestling.tr.x<-scale.mat(preytot.nestling.tr)
preytot.w1.tr.x<-scale.mat(preytot.w1.tr)


```

##D.1 first and second brood

```{r}
sink("D.1.jags")
cat("
    model {

    for (j in 1:2){
      alpha[j,1] ~ dnorm(1, 0.1)                                  #sex specific prior for alpha 1, Intercept number of eggs
      beta[j,1] ~ dunif(-10, 10)                                  #sex specific prior for beta 1, Intercept hatching success
      gamma[j,1] ~ dunif(-10, 10)                                 #sex specific prior for gamma 1, Intercept fledging success
      delta[j,1] ~ dnorm(0, 0.44)                                 #sex specific prior for delta 1, Intercept survival
      
      sigma.col.p[j] ~ dunif(0,10)                                #sex specific prior for precision coloration                           
      tau.col.p[j] = 1/(sigma.col.p[j] * sigma.col.p[j])          #sex specific prior for precisicion coloration
      
      for (k in 2:5){
      alpha[j,k] ~ dnorm(0, 0.1)                                  #sex specific prior for alphas, beta coefficients number of eggs
      }

      for (k in 2:7){
      beta[j,k] ~ dunif(-10, 10)                                  #sex specific prior for betas, beta coefficients hatching success
      }
      
      for (k in 2:7){
      gamma[j,k] ~ dunif(-10, 10)                                 #sex specific prior for gammas, beta coefficients fledging success
      }
      
      for (k in 2:12){
      delta[j,k] ~ dnorm(0, 0.1)                                  #sex specific prior for deltas, beta coefficients survival
      }
      
    }
    
    
    p[1,1] ~ dbeta(1,1)                                           #priors for sex and VHF specific recovery and recapture/resighting
    p[1,2] ~ dbeta(1,1)
    p[2,1] ~ dbeta(1,1)
    p[2,2] ~ dbeta(1,1)
    r[1,1] ~ dbeta(1,1)
    r[1,2] ~ dbeta(1,1)
    r[2,1] ~ dbeta(1,1)
    r[2,2] ~ dbeta(1,1)
    
    

    for (t in 1:(n.years-1)){                                       #priors for year specific precision of cort
      
      sigma.cort.r[t] ~ dgamma(1,1)
      tau.cort.r[t] = 1/(sigma.cort.r[t] * sigma.cort.r[t])
      sigma.cort.b[t] ~ dgamma(1,1)
      tau.cort.b[t] = 1/(sigma.cort.b[t] * sigma.cort.b[t])
      }


    for (i in 1:I){
    
      z[i,f[i]] <- 1
      
      for (t in f[i]:(n.years-1)){                                          #linear predictor for nr. eggs
        log(psi.e[i,t]) = alpha[sex[i],1] +
                          alpha[sex[i],2] * age.cat[i,t] +                  #age category, experienced or unexperienced
                          alpha[sex[i],3] * lay.d.x[i,t] +                  #laying date
                          alpha[sex[i],4] * preytot.play.pl.x[i,t] +        #prey index for plates one month before laying
                          alpha[sex[i],5] * preytot.play.tr.x[i,t]          #prey index for transects one month before laying
                                  
      }
      

      for (t in f[i]:(n.years-1)){                                          #linear predictor hatching success
        logit(rho.n[i,t]) = beta[sex[i],1] +
                            beta[sex[i],2] * age.cat[i,t] +                 #age category, experienced or unexperienced
                            beta[sex[i],3] * egg2.x[i,t] +                  #number of annually produced eggs
                            beta[sex[i],4] * lay.d.x[i,t] +                 #laying date
                            beta[sex[i],5] * BCIe.x[i,t] +                  #body condition after egg laying
                            beta[sex[i],6] * preytot.lay.pl.x[i,t] +        #prey index for plates at laying of first egg
                            beta[sex[i],7] * preytot.lay.tr.x[i,t]          #prey index for transects at laying of first egg
      }
      
      for (t in f[i]:(n.years-1)){                                          #linear predictor fledging success
        logit(rho.f[i,t]) = gamma[sex[i],1] +
                            gamma[sex[i],2] * age.cat[i,t] +                #age category, experienced or unexperienced
                            gamma[sex[i],3] * nestling2.x[i,t] +            #number of annually produced nestlings
                            gamma[sex[i],4] * lay.d.x[i,t] +                #laying date
                            gamma[sex[i],5] * BCIn.x[i,t] +                 #body condition during nestling phase
                            gamma[sex[i],6] * preytot.nestling.pl.x[i,t] +  #prey index for plates during nestling period
                            gamma[sex[i],7] * preytot.nestling.tr.x[i,t]    #prey index for transects during nestling period
      }
      
      

      for (t in (f[i]+1):n.years){
      

        logit(phi[i,t-1]) = delta[sex[i],1] +                               #linear predictor for survival
                            delta[sex[i],2] * age.cat[i,t-1] +
                            delta[sex[i],3] * BCIn.x[i,t-1] +               # body condition during nestling phase
                            delta[sex[i],4] * nestling2.x[i,t-1] +          # number of total nestlings produced
                            delta[sex[i],5] * fledge2.x[i,t-1] +            # number of fledglings produced
                            delta[sex[i],6] * cort.b.x[i,t-1] +             # cort baseline
                            delta[sex[i],7] * cort.r.x[i,t-1] +             # cort response
                            delta[sex[i],8] * col.p.x[i,t-1] +              # coloration, point diameter
                            delta[sex[i],9] * preytot.lay.pl.x[i,t-1] +     # prey index for plates during at laying from first egg
                            delta[sex[i],10] * preytot.lay.tr.x[i,t-1] +    # prey index for transects during at laying from first egg
                            delta[sex[i],11] * preytot.w1.pl.x[i,t-1] +     # prey index for plates during early winter (Nov, Dec)
                            delta[sex[i],12] * preytot.w1.tr.x[i,t-1]       # prey index for transects during early winter (Nov, Dec)
                            #delta[sex[i],] * snowdays5s.x[i,t-1]           # number of days with >5cm snow during oncoming winter
                            #delta[sex[i],] * tdlow5w.x[1,t-1]              # number of days with temp <5 degrees during winter
                            


        z[i,t] ~ dbern(z[i,t-1] * phi[i,t-1])                               # latent state  
        alive[i,t] ~ dbern(z[i,t] * p[sex[i], gps[i,t-1]])                  # the recapture probability
        dead[i,t] ~ dbern((z[i,t-1] - z[i,t]) * r[sex[i], gps[i,t-1]])      # the recovery probability
      }
      
      for (t in f[i]:(n.years-1)){                                          #sampling missing values in the data, with pre-definied parameters
        egg2[i,t] ~ dpois(psi.e[i,t])
        egg2.x[i,t]=(egg2[i,t]-egg2.mean)/egg2.sd
        nestling2[i,t] ~ dbin(rho.n[i,t], egg2[i,t])
        nestling2.x[i,t]=(nestling2[i,t]-nestling2.mean)/nestling2.sd
        fledge2[i,t] ~ dbin(rho.f[i,t], nestling2[i,t])
        fledge2.x[i,t]=(fledge2[i,t]-fledge2.mean)/fledge2.sd
        lay.d.x[i,t] ~ dnorm(mu.lay.d[t], tau.lay.d[t])
        col.p.x[i,t]=(col.p[i,t]-col.p.mean)/col.p.sd
        cort.r.x[i,t]=(cort.r[i,t]-cort.r.mean)/cort.r.sd
        cort.b.x[i,t]=(cort.b[i,t]-cort.b.mean)/cort.b.sd
        BCIe.x[i,t] ~ dnorm(mu.bcie[sex[i],t], tau.bcie[sex[i],t])
        BCIn.x[i,t] ~ dnorm(mu.bcin[sex[i],t], tau.bcin[sex[i],t])
        preytot.lay.pl.x[i,t] ~ dnorm(mu.lay.pl[t], tau.lay.pl[t])
        preytot.play.pl.x[i,t] ~ dnorm(mu.play.pl[t], tau.play.pl[t]) 
        preytot.nestling.pl.x[i,t] ~ dnorm(mu.nestling.pl[t], tau.nestling.pl[t])
        preytot.w1.pl.x[i,t] ~ dnorm(mu.w1.pl[t], tau.w1.pl[t])
        preytot.lay.tr.x[i,t] ~ dnorm(mu.lay.tr[t], tau.lay.tr[t])
        preytot.play.tr.x[i,t] ~ dnorm(mu.play.tr[t], tau.play.tr[t]) 
        preytot.nestling.tr.x[i,t] ~ dnorm(mu.nestling.tr[t], tau.nestling.tr[t])
        preytot.w1.tr.x[i,t] ~ dnorm(mu.w1.tr[t], tau.w1.tr[t])

      }
      
      for(t in (f[i]+1):(n.years-1)){
        col.p[i,t] ~ dnorm(col.p[i,t-1], tau.col.p[sex[i]])
        cort.r[i,t] ~ dnorm(cort.r[i,t-1], tau.cort.r[t])
        cort.b[i,t] ~ dnorm(cort.b[i,t-1], tau.cort.b[t])

      }

    }

# end model
  }
  ",fill = TRUE)
sink()

```

```{r}
jags.data <- list(alive = alive, dead = dead, sex = sex, n.years = ncol(alive), gps = gps, f = f, I = nrow(alive), z = z,
                  age.cat = age.cat,
                  egg2 = egg2, 
                    egg2.sd = sd(egg2, na.rm=T), 
                    egg2.mean = mean(egg2, na.rm=T),
                  nestling2 = nestling2, 
                    nestling2.sd = sd(nestling2, na.rm=T), 
                    nestling2.mean = mean(nestling2, na.rm=T),
                  fledge2 = fledge2, 
                    fledge2.sd = sd(fledge2, na.rm=T), 
                    fledge2.mean = mean(fledge2, na.rm=T),
                  col.p = col.p, 
                    col.p.sd = sd(col.p, na.rm=T),
                    col.p.mean = mean(col.p, na.rm=T),
                  cort.r = cort.r, 
                    cort.r.sd = sd(cort.r, na.rm=T), 
                    cort.r.mean = mean(cort.r, na.rm=T),
                  col = col, 
                    col.sd = sd(col, na.rm=T),
                    col.mean = mean(col, na.rm=T),
                  cort.b = cort.b, 
                    cort.b.sd = sd(cort.b, na.rm=T), 
                    cort.b.mean = mean(cort.b, na.rm=T),
                  lay.d.x=lay.d.x,
                    tau.lay.d=1/(colSds(lay.d.x, na.rm=T))^2, 
                    mu.lay.d=colMeans(lay.d.x, na.rm=T),
                  BCIe.x = BCIe.x, 
                    tau.bcie=rbind((1/(colSds(BCIe.x[sex==1,], na.rm=T))^2),(1/(colSds(BCIe.x[sex==2,], na.rm=T))^2)), 
                    mu.bcie=rbind(colMeans(BCIe.x[sex==1,], na.rm=T),colMeans(BCIe.x[sex==2,], na.rm=T)),
                  BCIn.x = BCIn.x, 
                    tau.bcin=rbind((1/(colSds(BCIn.x[sex==1,], na.rm=T))^2),(1/(colSds(BCIn.x[sex==2,], na.rm=T))^2)), 
                    mu.bcin=rbind(colMeans(BCIn.x[sex==1,], na.rm=T),colMeans(BCIn.x[sex==2,], na.rm=T)),
                  preytot.play.pl.x = preytot.play.pl.x, 
                    tau.play.pl=1/(colSds(preytot.play.pl.x, na.rm=T))^2, 
                    mu.play.pl=colMeans(preytot.play.pl.x, na.rm=T),
                  preytot.lay.pl.x = preytot.lay.pl.x, 
                    tau.lay.pl=1/(colSds(preytot.lay.pl.x, na.rm=T))^2, 
                    mu.lay.pl=colMeans(preytot.lay.pl.x, na.rm=T),
                  preytot.nestling.pl.x = preytot.nestling.pl.x,
                    tau.nestling.pl=1/(colSds(preytot.nestling.pl.x, na.rm=T))^2, 
                    mu.nestling.pl=colMeans(preytot.nestling.pl.x, na.rm=T),
                  preytot.w1.pl.x = preytot.w1.pl.x, 
                    tau.w1.pl=1/(colSds(preytot.w1.pl.x, na.rm=T))^2, 
                    mu.w1.pl=colMeans(preytot.w1.pl.x, na.rm=T),
                  preytot.play.tr.x = preytot.play.tr.x, 
                    tau.play.tr=1/(colSds(preytot.play.tr.x, na.rm=T))^2, 
                    mu.play.tr=colMeans(preytot.play.tr.x, na.rm=T),
                  preytot.lay.tr.x = preytot.lay.tr.x, 
                    tau.lay.tr=1/(colSds(preytot.lay.tr.x, na.rm=T))^2, 
                    mu.lay.tr=colMeans(preytot.lay.tr.x, na.rm=T),
                  preytot.nestling.tr.x = preytot.nestling.tr.x,
                    tau.nestling.tr=1/(colSds(preytot.nestling.tr.x, na.rm=T))^2, 
                    mu.nestling.tr=colMeans(preytot.nestling.tr.x, na.rm=T),
                  preytot.w1.tr.x = preytot.w1.tr.x, 
                    tau.w1.tr=1/(colSds(preytot.w1.tr.x, na.rm=T))^2, 
                    mu.w1.tr=colMeans(preytot.w1.tr.x, na.rm=T)
                  )
                  

parameters <- c("p", "r", "alpha", "beta", "gamma", "delta", "m.rho.n")


inits <- function(){list()}

nc <- 4
nt <- 25
ni <- 25000
nb <- 1000


D.1 <- jags(jags.data, inits, parameters, "D.1.jags", parallel = T,
           n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb)


```

print model estimates
```{r}

print(D.1, digits=3)

```

